Synthesizing Symmetric Lenses

### 2018-05-29 papers/SymmetricLenses/main.tex
add data

@@

% Creates a display mode for code in sans serif font
\lstnewenvironment{sflisting}[1][]
  {\lstset{%
    mathescape,
    basicstyle=\small\sffamily,
    aboveskip=5pt,
    belowskip=5pt,
    columns=flexible,
    frame=,
    xleftmargin=1em,#1}\sansmath}
  {}
% end

@@

\begin{abstract}
  Similar data is oftentimes represented and stored in different formats. Users
  commonly need to synchronize the data shared between formats, where edits to
  the data in one format gets propagated to the data in the other format.
  Manually writing these data synchronizers is tedious and error prone. To
  alleviate these difficulties, tools have been developed that automatically
  generate these synchronization functions by generating bidirectional lens
  programs from data format descriptors~\cite{optician,quotient-optician}.
  However, these tools only work when the two formats describe exactly the same
  data; it assumes that all important information present in one format is
  present in the other. This is often not the case.

  In this work, we automatically synthesize conversion functions between formats
  that each may contain information the other lacks. We define \emph{simple
    symmetric lenses}, a class of bidirectional programs for expressing these
  types of conversion functions, and we specify a language that expresses simple
  symmetric lenses that operate on string formats. We develop a type-directed
  synthesis algorithm for generating lenses in this language from regular
  expressions describing the data formats. A significant technical challenge in
  this work arises from the fact that our target language permits
  synchronization between formats that don't have exactly the same data -- data
  in one format need not be present in the other format. This gives the
  synthesis algorithm much more freedom, as data can be completely
  unsynchronized (edits in one format are never propagated), completely
  synchronized (the formats are in bijective correspondence), or anywhere in
  between. Under the assumption that users typically want to synchronize as much
  data as possible, we provide an information theoretic metric for how much data
  is left unsynchronized based on the theory of \emph{stochastic regular
    expressions}. Furthermore, we use this metric to develop a greedy synthesis
  algorithm that tries to minimize unsynchronized data. We experimentally
  demonstrate that despite addressing a much harder problem, our extension to
  Optician still maintains synthesis times competitive to Optician on Optician's
  bijective benchmark suite, while generalizing to benchmarks synchronizing data
  not in bijective correspondence.
\end{abstract}

@@

Unfortunately, writing such synchronizers is tedious and error prone.
Bidirectional languages ease the difficulty of writing such synchronizers, as a
single, well-typed program expresses multiple conversion functions, and
guarantee these functions satisfy round-tripping laws. However, bidirectional
languages are difficult to program in as well -- often requiring users to
permute data in fiddly ways while satisfying the type system's complex
unambiguity constraints.

A better approach is to synthesize such conversion functions. Recent research
has gone into automatically synthesizing such conversion functions from
input-output examples and formal descriptions of the data
formats~\cite{optician,quotient-optician}. However, this work assumes that all
data fields present in one format are present in the other, but this is not
always true. Oftentimes, fields present in one format are not present in the
other: the commands scheduled to run on Linux machines are typically different
than the commands scheduled to run on Windows machines; updated APIs can expose
new information, or omit legacy information; ICS files contain a URL field that
Google Calendar CSV files do not expose. We aim to synthesize transformations
where the data fields are not necessarily in bijective correspondence, where
each data format can contain important information not present in the other
format.

We follow a similar approach to existing work on synthesizing conversion
functions; instead of synthesizing these functions individually, we synthesize
bidirectional programs. To this end, we develop a new formulation of
bidirectional programs, \emph{simple symmetric lenses}, which permit
synchronizations between formats not in bijective correspondence. We develop a
language for expressing simple symmetric lenses that operate on string data.
Terms in this language are typed between regular expressions, where the regular
expressions describe the synchronized data formats.

Our synthesis algorithm takes as input a suite of input-output examples and two
regular expressions describing the data formats. Our synthesis algorithm uses
type-directed synthesis on the two regular expressions to synthesize a lens that
is well-typed between the provided regular expressions and satisfies the
provided input-output examples.

A primary novel difficulty is that, unlike previous work where there are
relatively few well-typed lenses, there are a relatively many well-typed simple
symmetric lenses. As symmetric lenses permit data to be present in only one
format, for any data component there is now a choice: is this data only relevant
to this format and as such should be left unsynchronized, or is this data
present in both formats and as such should be synchronized.

@@

As we wish to synchronize as much data as we can, our cost model gives higher
cost to those with more data is left unsynchronized. In particular, the cost of
a lens is the expected number of bits required to recover a string in one format
when a synchronized string in the other format is provided. In this model,
bijective lenses have zero cost, and as more information is left unsychronized,
the cost is increased.

Our model for this is based on \emph{stochastic regular expressions}~\cite{?}.
We represent a probability distribution for strings by a stochastic regular
expression, and find a way to syntactically infer the expected information
content (or \emph{entropy}) of an unambiguous regular expression. We use this
entropy to calculate the expected information content to recover a string, when
we are given a string it is synchronized to.

@@

We develop an algorithm that synthesizes a low-cost lens, shown in
Figure~\ref{fig:high-level-algorithm}. Our algorithm consists of two
communicating synthesizers, one that proposes candidate stochastic regular
expression pairs that it thinks will align well, and one that greedily searches
for a lens between those two regular expressions that minimizes cost and
involves no rewrites. When the first algorithm thinks it has found the best
possible lens, it returns that lens.

@@

In summary, our contributions are:
\begin{enumerate}
\item We develop \emph{simple symmetric lenses} a new formulation for
  bidirectional programs that can project data from both sides~(\S\ref{sec:ssl}),
  and describe a language that expresses such lenses on string
  data~(\S\ref{sec:ssl-string}).
\item We develop a means to traverse stochastic regular expression equivalences,
  while preserving probability distributions and
  languages~(\S\ref{sec:stoch-rx}).
\item We develop stochastic DNF regular expressions and DNF lenses. We develop a
  sound algorithm for converting stochastic regular expressions to stochastic
  DNF regular expressions~(\S\ref{sec:dnf-regex-lens}), and a
  information-theoretic cost metric on DNF
  lenses~(\S\ref{sec:symmetric-dnf-lenses}).
\item We extended Boomerang with our simple symmetric lens combinators. We
  updated the Optician synthesis system to synthesize symmetric lenses with our
  new algorithm.
\item We evaluate our synthesis algorithm on Opticians benchmark suite as well
  as new, symmetric benchmarks, and find that we lose very little synthesis
  time, while gaining a large amount of expressivity (\S\ref{sec:evaluation}).
\end{enumerate}
% end introduction

% begin extended example
\section{Extended Example}

\begin{figure}
  % https://tex.stackexchange.com/questions/168741/showing-two-listings-in-a-table-side-by-side
  \setbox0=\hbox{%
    \begin{minipage}{3in}
\begin{lstlisting}[
tabsize=1,
language={xml},
]
0,20,40 cp -r /Library/ "$\$$(date)"

@@

  \setbox0=\hbox{%
    \begin{minipage}{3in}
\begin{lstlisting}[
tabsize=1,
]
<Tasks>
 <Task>
  <Command>c</Command>
  <Time>0</Time><Time>20</Time><Time>40</Time>
 </Task>
 <Task>
  <Command>c</Command>
  <AllTimes/>
 </Task>
<Tasks>

{
  Title: Return 400 on
    bad PUT request,
  Type: Bug,
  AssignedTo: J Smith,
}
\end{lstlisting}
    \end{minipage}
  }
  \savestack{\listingB}{\box0}
  \centering
  \resizebox{\columnwidth}{!}{%
    \begin{tabular}{|>{\columncolor{lightyellow}}c|>{\columncolor{lightgreen}}c|}
      \hline
      Legacy Work Item Representation & Modern Work Item Representation \\
      \hline
      \listingA & \listingB \\
      \hline
    \end{tabular}
  }
  \caption{Hello there this is the thing}
  \label{fig:minimized-representations}
\end{figure}

We motivate our problem with an extended example.  Consider the previously
described example of keeping Linux and Windows scheduled jobs synchronized.
In this example, we describe keeping synchronized an minified version of
crontab and the task scheduler.  These are each formats specifying when to run
recurring tasks.  For simplicity, we are using an minified version of these
configuration files where we can schedule tasks to occur at fixed intervals,
hourly, daily, or weekly.  Each task consists of a command to be run.

In this example, we wish for the jobs to be synchronized across our windows and
linux devices. However, the commands that need to be run are typically different
for each device. Paths are different in windows and Linux, and while there is
typically a Windows equvialent for Linux commands, that equivalent command is
not always named the same thing.

To synchronize these files, we wish to synchronize how often the commands are
run, but not synchronize the names of the commands. We consider 4 functions to
be necessary to synchronize these two files: a function that makes a new cron
file from a newly created task scheduler, a function that makes a new task
scheduler file from a new cron file, a function that merges edits to the cron
file to the synchronized task scheduler file, and a function that merges edits
on the task scheduler file to the cron file. These are the only four functions
we need to synchronize these two files, so these four functions comprise a
simple symmetric lens between cron files and task scheduler files.  More
formally, a simple symmetric lens between $X$ and $Y$ comprise the following
four functions:
\begin{enumerate}
\item $\CreateR \OfType X \rightarrow Y$
\item $\CreateL \OfType Y \rightarrow X$
\item $\PutR \OfType X \rightarrow Y \rightarrow Y$
\item $\PutL \OfType Y \rightarrow X \rightarrow X$
\end{enumerate}

Furthermore, we expect these functions to satisfy certain properties.  For
example, if two files are synchronized by one of these functions, using another
function to resynchronize it will not change the result.  These are viewed as
the ``round-tripping'' laws:
\vspace*{-1em}
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}

\vspace*{-1em}
We wish 
% end extended example

>>>>

\lstset{ language=Caml, basicstyle=\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkpurple}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={rep,iterate,synth,collapse,perm,squash,normalize,using,ins,del,lens,let,get,put,rquot,lquot,id,swap,concat,or,disc,merge_left,merge_right,const},
emphstyle=\upshape\sffamily\color{dkpurple}, 
columns=fullflexible,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
stringstyle=\sffamily\color{dkblue},
}
\makeatletter
     \let\lst@oldvisiblespace\lst@visiblespace
     \def\lst@visiblespace{\,\lst@oldvisiblespace\,}
\makeatother

@@

% Creates a display mode for code in sans serif font
% end

@@

\begin{abstract}
  Users commonly need to synchronize the data shared between formats, so edits
  to the data in one format gets propagated to the data in the other format.
  Manually writing the functions for these data synchronizers is tedious and
  error prone. Instead of manually writing these synchronizers, bidirectional
  languages can be used to express \emph{lenses}, groups of synchronization
  functions with guarantees about round-trip behavior. While these bidirectional
  languages help make programming these synthesizers less error-prone, they
  remain difficult to program in. Synthesizing bidirectional languages allows
  users to use lenses without manually programming them. Current
  synthesizers work only on the simplest types of lenses, when the two formats
  describe exactly the same data; this is often not the case.

  Oftentimes, each format may contain information the other lacks. Such
  synchronization tasks are described by one of the richest classes of lenses,
  symmetric lenses. However, symmetric lenses in their current formulation
  require external state in the form of a \emph{complement}. This statefulness
  makes symmetric lenses both difficult to use and difficult to synthesize.

  In this work, we reformulate symmetric lenses as \emph{simple symmetric
    lenses}. While this formulation is less expressive than symmetric lenses, it
  expresses many symmetric lenses while requiring \emph{no external state}. We
  provide combinators for simple symmetric lenses on string data, and integrate
  these combinators into Boomerang while providing full backwards compatibility.

  We develop a type-directed synthesis algorithm for generating lenses in this
  language from regular expressions describing the data formats. A significant
  technical challenge in this work arises from the fact that our target language
  permits synchronization between formats that don't have exactly the same data
  -- data in one format need not be present in the other format. This gives the
  synthesis algorithm much more freedom, as data can be completely
  unsynchronized (edits in one format are never propagated), completely
  synchronized (the formats are in bijective correspondence), or anywhere in
  between. Under the assumption that users typically want to synchronize as much
  data as possible, we provide an information theoretic metric for how much data
  is left unsynchronized based on the theory of \emph{stochastic regular
    expressions}. Furthermore, we use this metric to develop a greedy synthesis
  algorithm that tries to minimize unsynchronized data. We experimentally
  demonstrate that our algorithm can efficiently synthesize simple symmetric
  lenses on a variety of benchmark suites consisting of synchronizing file
  formats from configuration files, application-specific data storage files, and
  data cleaning tasks.

\end{abstract}

@@

Unfortunately, writing such synchronizers is tedious and error prone.
Bidirectional languages ease the difficulty of writing such synchronizers, as a
single, well-typed program expresses a \emph{lens}, a group of conversion
functions guaranteed to satisfy round-tripping laws. However, bidirectional
languages are difficult to program in as well, often requiring users to permute
data in fiddly ways while satisfying the type system's complex unambiguity
constraints.

A better approach is to synthesize such conversion functions. Recent research
has gone into Optician, a tool for automatically synthesizing such conversion
functions from input-output examples and formal descriptions of the data
formats~\cite{optician,quotient-optician}. However, this work assumes that all
data fields present in one format are present in the other, but this is not
always true. Oftentimes, fields present in one format are not present in the
other: the commands scheduled to run on Linux machines are typically different
than the commands scheduled to run on Windows machines; updated APIs can expose
new information, or omit legacy information; ICS files contain a URL field that
Google Calendar CSV files do not expose.

\emph{Symmetric lenses} formalize such conversion functions. However, these
symmetric lenses use external state for synchronization, in the form of a
\emph{complement}. This makes them difficult to use in real-world
synchronization tasks, as we must maintain the files we synchronize as well as
maintaining the complements. The complement further complicates synthesis, as
input-output examples are no longer sufficient, instead edit sequences must be
provided.

To get around this, we reformulate symmetric lenses as \emph{simple symmetric
  lenses}. This reformulation does not use a complement, and only uses the files
that we aim to synchronize as state. We identify a property,
\emph{forgetfulness}, where if a symmetric lens can be written as a forgetful
symmetric lens, it can also be written a simple symmetric lens. We provide
combinators for building simple symmetric lenses on string data, and integrate
these combinators with Boomerang, a language for bidirectional programming on
strings. In Boomerang, and our extension to it, lens expressions are typed
between regular expressions, where the regular expressions describe the
synchronized data formats.

We enhance Optician with simple symmetric lens synthesis capabilities. This
extension synthesizes symmetric lenses from a suite of input-output examples and
a pair of regular expressions describing the data formats. The synthesis
algorithm uses type-directed synthesis on the two regular expressions to
synthesize a lens that is well-typed between the provided regular expressions
and satisfies the provided input-output examples.

A primary difficulty is that there are a relatively many well-typed simple
symmetric lenses. As symmetric lenses permit data to be present in only one
format, for any data component there is now a choice: is this data only relevant
to this format and as such should be left unsynchronized, or is this data
present in both formats and as such should be synchronized.

@@

In particular, the cost of a lens is the expected number of bits required to
recover a string in one format when a synchronized string in the other format is
provided. In this model, bijective lenses have zero cost, and as more
information is left unsychronized, the cost is increased.

Our model for this is based on \emph{stochastic regular expressions}~\cite{?}. A
probability distribution for strings is represented by a stochastic regular
expression, and we develop a way to syntactically infer the expected information
content (or \emph{entropy}) of an unambiguous regular expression. This entropy
is used to calculate the expected information content to recover a string, when
given a string it is synchronized to.

@@

We develop an algorithm that synthesizes a low-cost lens, shown in
Figure~\ref{fig:high-level-algorithm}. This algorithm consists of two
communicating synthesizers, one that proposes candidate stochastic regular
expression pairs that it thinks will align well, and one that greedily searches
for a lens between those two regular expressions that minimizes cost and
involves no rewrites. When the first algorithm thinks it has found the best
possible lens, it returns that lens.

@@

In summary, our contributions are:
\begin{enumerate}
\item We develop \emph{simple symmetric lenses} a new formulation for
  bidirectional programs that can project data from both sides~(\S\ref{sec:ssl}),
  and describe a language that expresses such lenses on string
  data~(\S\ref{sec:ssl-string}).
\item We develop a means to traverse stochastic regular expression equivalences,
  while preserving probability distributions and
  languages~(\S\ref{sec:stoch-rx}).
\item We develop stochastic DNF regular expressions and DNF lenses. We develop a
  sound algorithm for converting stochastic regular expressions to stochastic
  DNF regular expressions~(\S\ref{sec:dnf-regex-lens}), and a
  information-theoretic cost metric on DNF
  lenses~(\S\ref{sec:symmetric-dnf-lenses}).
\item We extended Boomerang with our simple symmetric lens combinators. We
  updated the Optician synthesis system to synthesize symmetric lenses with our
  new algorithm.
\item We evaluate our synthesis algorithm on Opticians benchmark suite as well
  as new, symmetric benchmarks, and find that we lose very little synthesis
  time, while being significantly more expressive (\S\ref{sec:evaluation}).
\end{enumerate}
% end introduction

% begin overview
\section{Overview}
\label{sec:overview}

\begin{figure}
  % https://tex.stackexchange.com/questions/168741/showing-two-listings-in-a-table-side-by-side
  \setbox0=\hbox{%
    \begin{minipage}{3.5in}
\begin{lstlisting}[
stringstyle=\upshape\sffamily
]
0,20,40 cp -r /Library/ "$\$$(date)"

@@

  \setbox0=\hbox{%
    \begin{minipage}{3.5in}
\begin{lstlisting}[
stringstyle=\upshape\sffamily
]
<Tasks>
 <Task>
  <Command>c</Command>
  <Time>0</Time><Time>20</Time> <Time>40</Time>
 </Task>
 <Task>
  <Command>c</Command>
  <AllTimes/>
 </Task>
<Tasks>
\end{lstlisting}
    \end{minipage}
  }
  \savestack{\listingB}{\box0}

  \setbox0=\hbox{%
    \begin{minipage}{3.5in}
\begin{lstlisting}
let minute_list = minute . ("," . minute)$^*$ in
let times = "*" | minute_list in
let line = times . " " . command in
let crontab = "" | line . ("\n\n" line)*
\end{lstlisting}
    \end{minipage}
  }
  \savestack{\listingC}{\box0}

  \setbox0=\hbox{%
    \begin{minipage}{3.5in}
\begin{lstlisting}
let minute_xml = "<Time>" . minute . "</Time>" in
let minutes_xml = minute_xml . minute_xml$^*$ in
let xml_time = "<AllTimes/>" | minutes_xml in
let command_xml = 
  "<Command>" . command . "</Command>"
in
let xml_task = command_xml . xml_time in
let scheduled_job = "<Tasks>" . xml_task$^*$ . "</Tasks>"
\end{lstlisting}
    \end{minipage}
  }
  \savestack{\listingD}{\box0}

  \centering
  \resizebox{\columnwidth}{!}{%
    \begin{tabular}{|>{\columncolor{lightyellow}}c|>{\columncolor{lightgreen}}c|}
      \hline
      crontab & scheduled.job \\
      \hline
      \listingA & \listingB \\
      \hline
      \listingC & \listingD \\
      \hline
    \end{tabular}
  }
  \caption{An example crontab and Scheduled Tasks file, and regular expressions
    describing the formats. For simplicity, these are modified versions of the
    real formats.}
  \label{fig:minimized-representations}
\end{figure}

We motivate our problem with an extended example. Consider the previously
described example of keeping Linux and Windows scheduled jobs synchronized. In
this example, we describe keeping synchronized a crontab (the file containing
cron jobs) and a Scheduled Tasks file, scheduled.job. These are each formats
specifying when to run recurring tasks. For simplicity of description, we use a
minified version of each of these, with examples and regular expressions shown
in Figure~\ref{fig:minimized-representations}.

For task is set in these files at specific times each hour. In both these
formats, users can specify specific minutes they want to run the program during,
or they can specify they want to run the program at every minute.

In this example, we wish for the jobs to be synchronized across our windows and
linux devices. However, the commands that need to be run are typically different
for each device. Paths are different in Windows and Linux, and while there is
typically a Windows equvialent for Linux commands, that equivalent command is
not always named the same thing.  So, while we want to be able to add commands
and edit how often they are run, we don't want to synchronize which commands are
run themselves.

To synchronize these files, a few functions are needed. Two \emph{creates} are
needed; if a crontab is initialized, a synchronized Scheduled Tasks file is also
created, and vice versa. Furthermore, we require \emph{puts}, functions that
take an updated crontab and an old Scheduled Tasks file, and output a new
Scheduled Tasks file with the information synchronized to the updated crontab.
To have a simple symmetric lens is to have these required synchronization
functions. Formally, a simple symmetric lens between $X$ and $Y$ comprises 4
functions:
\begin{enumerate}
\item $\CreateR \OfType X \rightarrow Y$
\item $\CreateL \OfType Y \rightarrow X$
\item $\PutR \OfType X \rightarrow Y \rightarrow Y$
\item $\PutL \OfType Y \rightarrow X \rightarrow X$
\end{enumerate}
Furthermore, if a file is not edited, but is then resynchronized using a put
function, no data should be changed. To be a lens, one must satisfy the
following ``round-tripping'' laws that guarantee identity edits won't change the
file data.
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}

For synchronization, when the original crontab or Scheduled Tasks file is
created, a synchronized file of the other form is created from it with
$\CreateR$ or $\CreateL$ (depending on which file was initially created). On
future edits, changes are made with the puts, a \PutR would propagate the shared
information from the crontab, and knit it together with the information specific
to the Scheduled Tasks file.

\subsection{Programming Symmetric Lenses} We now build up a symmetric lens by
example. Our symmetric lens expressions are typed between regular expression
pairs, if \lstinline{$\Lens$ : $\Regex \Leftrightarrow \RegexAlt$}, then $\Lens$
encodes a lens between $\LanguageOf{\Regex}$ and $\LanguageOf{\RegexAlt}$. At
the end, we want our symmetric lenses to by typed between \lstinline{crontab}
and \lstinline{scheduled_job}.

The identity lens takes, as argument, a regular expression, and performs the
identity function on strings in that regular expression. The identity lens fully
synchronizes the data formats: when one data format is updated, the other format
gets the exact same value. In our example, when a job is told to run at a
specific time in crontab, it is also expected to run at that same time in the
Scheduled Jobs list, so the identity lens would propagate that information well.
%
\begin{lstlisting}
id(minute) : minute $\Leftrightarrow$ minute
\end{lstlisting}
%

Where the identity lens synchronizes data, the disconnect lens does the
opposite: it leaves data unsynchronized. The disconnect lens takes 4 arguments:
two regular expressions and two strings. On creates, because there is no
association between data in the first format and data in the second format,
designated default strings $\String$ and $\StringAlt$ are put into the other
format. In our extended example, the Linux commands should not by synchronized
with the Windows commands, so a disconnect lens would be used. We want our
default behavior to be failures, under the assumption this failure will raise a
warning, in case somebody forgets to find the analogous command of the other
operating system.
%
\begin{lstlisting}
disc(command,command_xml,"false","<Command>SET ERRORLEVEL=1</Command>") 
  : command $\Leftrightarrow$ command_xml
\end{lstlisting}
%
When building a disconnect lens between two strings, one of which is merely the
empty string, the lenses \lstinline{ins} and \lstinline{del} can instead be
used, where \lstinline{ins($t$) = disc("",$t$,"",$t$)} and
\lstinline{del($s$) = disc($s$,"",$s$,"")}.

Finally, there are a number of lenses that compositionally build up new lenses
from old ones. For example, data can appear in one form or another, a program
can run during a specific set of minutes, or that program can run the every
minute. If \lstinline{minutes_map} and \lstinline{alltimes_map} are lenses that
synchronize those settings in each format, then those lenses can be composed
with an \OrLens lens.
%
\begin{lstlisting}
or(minutes_map,alltimes_map) : times $\Leftrightarrow$ xml_time
\end{lstlisting}

Data can also be concatenated, and so lens mappings can be concatenated.
However, data concatenations can only happen in a number of ways, as data can
appear in the same order, and data can be reordered. For example, in the
crontab, time data appears before command data, whereas in the Scheduled Jobs
file, command data comes before time data. If \lstinline{command_map} provides a
lens between commands in each format, and \lstinline{time_map} provides a lens
between time data in each format, those lenses can be composed with a \SwapLens
lens.
%
\begin{lstlisting}
swap(command_map,time_map) : line $\Leftrightarrow$ xml_task
\end{lstlisting}
%
Two lenses, $\Lens_1$ and $\Lens_2$, can be concatenated as
\lstinline{concat($\Lens_1$,$\Lens_2$)} or as the infix style
\lstinline{$\Lens_1$ . $\Lens_2$} for data that is not reordered.

Lastly, the \IterateLens lens synchronizes repeated data, synchronizing each
element of the iteration, and synchronizing the number of iterations. For
example, if \lstinline{full_task_map} synchronizes crontab lines prepended by
two newlines and xml for a single task, iterating that lens builds a lens that
synchronizes a list of tasks.
%
\begin{lstlisting}
iterate(full_task_map) : ("\n\n" . line)$^*$ $\Leftrightarrow$ xml_task$^*$
\end{lstlisting}
%

\begin{figure}
\begin{lstlisting}
let full_minute_map = ins("<Time>") . id(minute) . ins("</Time>") in
let comma_minute_map = del(",") . full_minute_map in
let minutes_map = full_minute_map . iterate(comma_minute_map) in
let alltimes_map = del("*") . ins("<AllTimes/>") in
let times_map = or(minutes_map,alltimes_map) in

let command_core = 
  disc(command,command_xml,"false","<Command>SET ERRORLEVEL=1</Command>")
in
let command_map = 
  ins("<Command>") . command_core . ins("</Command>")) . del(" ") 
in

let task_map = swap(command_map,time_map) in
let full_task_map = 
  ins("<Task>") . del("\n\n") . task_map . ins("</Task>") 
in
let repeated_task_map = iterate(full_task_map)
let nonempty_task_map = ins("<Tasks>") . repeated_task_map . ins("</Tasks>")

or(ins("<Tasks></Tasks>"),nonempty_task_map) : crontab $\Leftrightarrow$ scheduled_job
\end{lstlisting}
  \caption{A lens, written in our language for building simple symmetric lenses
    on string data, that synchronizes crontab files and Scheduled Task files.
    Note that we can annotate our lenses with types, to confirm the lens
    operates on the languages of the regular expressions.}
  \label{fig:example_lens}
\end{figure}

Using these combinators, we can build up a lens that synchronizes crontabs and
Scheduled Task files, shown in Figure~\ref{fig:example_lens}. This lens uses one
typing rule not previously mentioned: type equivalence.  While
\lstinline{scheduled_job} is a equivalent to a lens with an outermost
disjunction, it is not syntactically equivalent to one.  Without type
equivalence, the provided lens wouldn't actually be well-typed, but type
equivalence states that if a lens is typed between a pair of regular
expressions, it is also typed between an equivalent pair.

Note that, while the programmed lens should be quite simple, it was relatively
difficult to write. We had to provide a lot of scaffolding to map between the
desired formats, even though the idea was quite simple: synchronize the times
but leave the commands unsynchronized. Furthermore, this is inherent to manually
writing these synchronization functions, as a programmer must describe which
components are synchronized with which, and which components are scaffolding.

\subsection{Synthesizing Symmetric Lenses}
To alleviate these difficulties, we wish to synthesize symmetric lenses from
types and examples. Given a pair of regular expressions and a set of
input-output examples, we want to synthesize a symmetric lens expression that is
typed between the regular expression pair, and satisfies the input/output
examples. For example, we wish to synthesize a lens from \lstinline{crontab} and
\lstinline{scheduled_job}, using the strings in
Figure~\ref{fig:minimized-representations} as input-output example for \CreateR.

However, our simple symmetric lens combinators permit many well-typed programs
between two regular expressions. For example, one possible lens between
\lstinline{crontab} and \lstinline{scheduled_job} is the one in
Figure~\ref{fig:example_lens}, but the lens
%
\begin{lstlisting}
disc(crontab,scheduled_job,rep(crontab),rep(scheduled_job))
\end{lstlisting}
%
is also a valid lens (where \lstinline{rep($\Regex$)} is a string in the
language of $\Regex$). Furthermore, adding the synchronized files from
Figure~\ref{fig:minimized-representations} as an input-output example helps
little, but the default strings given to \Disconnect merely need to be those
strings and the lens still satisfies the examples. While the disconnect lens is
prohibited by two examples, or by a carefully chosen input-output example on
\PutR or \PutL, this problem continues presenting itself for the sublenses, and
is highlighted even more on larger formats. While we could merely ask
programmers to provide large, tight specifications for their synthesis tasks, we
would prefer the algorithm to choose the right lens automatically. Instead of
merely finding \emph{any} satisfying lens, we wish to synthesize the \emph{best}
satisfying lens.

But what is the best satisfying lens?  We think the best satisfying lens is the
one that synchronizes as much data as is permitted by the type system and the
examples.  More formally, we wish to minimize the expected number of bits
required to recover data in one format from the other.  The expected number of
bits is a well known concept in information theory, the expected number of bits
of a data source is the \emph{entropy} of that data source.

The entropy of a data source is based on the probability distribution of that
data source. Given a set $S$, and a probability distribution over that set $P
\OfType S \rightarrow \Reals$, the entropy of $(S,P)$ is $-\Sigma_{s\in
  S}P(s)\Log_2P(s)$. While we have ways of expressing sets, we do not have
probability distributions. In this work, we use \emph{stochastic regular
  expressions}, regular expressions annotated with probabilities that express
languages and probability distributions over those languages. In this work, we
guess a reasonable probability distribution from the syntax, though if users
manually provided stochastic regular expressions, or large suite of data in the
formats through which we could learn the annotations, we could instead use that.

With sets and probability distributions, we can find the entropy of the
stochastic regular expression, given another stochastic regular expression, and
a lens that shows which strings are synchronized with which. We aim to find the
lens that minimizes that entropy.  Now that we have a notion of best, we need a
way to search for that best lens.

Prior work on lens synthesis works with two communicating synthesizers, one that
proposes regular expressions, semantically but not syntactically equivalent to
each other, and another that finds lenses between them, if there are any lenses
between them that don't use a type equivalence rule involving star equivalences.
We utilize this same framework, except instead of having the second synthesizer
find \emph{any} lens between the types, \GreedySynth instead uses a greedy
algorithm to try finding the \emph{best} lens between the types.

This poses a new problem: unless we find a lens with no cost, further search may
find a better lens. While we may have found a great lens between two regular
expressions, a different regular expression pair may provide an even better
lens. There may be a great lens between \lstinline{crontab} and
\lstinline{scheduled_job} that uses no star equivalences, but the lens shown in
Figure~\ref{fig:example_lens} does. However, there is a tension, we want to find
a good lens, but we also want to give quick feedback to the user. After finding
the lens in Figure~\ref{fig:example_lens}, the algorithm shouldn't spend much
longer searching for better lenses.

We address this by keeping running track of the best lens found thus far. Before
proposing new types, the regular expression generator, \RXSearch, evaluates
whether it should continue proposing regular expression pairs, or whether it is
content with the current best. For a given regular expression pair, \RXSearch
looks at the number of rewrite rules performed on the pair, and identifies how
many pairs with involving the same number of rewrites may be proposed. If there
are a large number of pairs that can be proposed with that many rewrites, or the
current lens is very well synchronized, \RXSearch does not propose that pairs.
In this way, \RXSearch will continue searching for better lenses, but does not
tackle huge search spaces unless it needs to. In this way, the algorithm can
provide lenses quickly to the users, which users can refine with additional
examples.
% end overview

% begin ssl
\section{Simple Symmetric String Lenses}
\label{sec:ssl}
We now give a formal description of all the combinators in our symmetric lens
language. These lens terms are typed between pairs of regular expressions: the
typing judgment $\Lens \OfType \Regex \Leftrightarrow \RegexAlt$ means that
$\Lens$ is a well typed lens between $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$.

\begin{center}
  \begin{tabular}{l@{\ }l@{\ }l@{\ }l}
    % REGEX
    \Lens{} & \GEq{} & & $\IdentityLensOf{\Regex}$\\
            & & \GBar{} & $\DisconnectOf{\Regex}{\RegexAlt}{\String}{\StringAlt}$ \\
            & & \GBar{} & $\IterateLensOf{\Lens}$ \\
            & & \GBar{} & $\ConcatLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\SwapLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\OrLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\ComposeLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\MergeROf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\MergeLOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\InvertOf{\Lens}$ \\
  \end{tabular}
\end{center}

We have already described over a number of these lenses in \S\ref{sec:overview}.
The compose lens $\Lens_1 ; \Lens_2$ composes two lenses sequentially, it
transforms from one data format to the other by first transforming to an
intermediate format, shared as the target of $\Lens_1$ and the source of
$\Lens_2$. The \MergeR lens combines two sublenses which share the same target.
Which lens gets used depends on what data is in the left-hand format, if it is
in the domain of $\Lens_1$ then $\Lens_1$ gets used, and similarly for
$\Lens_2$.  The \MergeL lens is symmetric to \MergeR, the two sublenses must
share the same source. Finally, $\InvertOf{\Lens}$ inverts a lens, \CreateR
becomes \CreateL and \PutR becomes \PutL, and vice-versa.

We now provide the typing derivations and semantics for these lenses. For
simplicity, we define the typing derivations and semantics only over unambiguous
regular expressions, even though some lenses can be defined over ambiguous
regular expressions (for example, the identity lens doesn't need any ambiguity
constraints).

\[
  \centering
  \inferrule*
  {
  }
  {
    \IdentityLensOf{\Regex} \OfType \Regex \Leftrightarrow \Regex
  }
\]
\begin{center}
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App s$ & = & $s$\\
    $\CreateL{} \App s$ & = & $s$\\
    $\PutR{} \App s_1 \App s_2$ & = & $s_1$\\
    $\PutL{} \App s_1 \App s_2$ & = & $s_1$
  \end{tabular}
\end{center}

Note that the identity lens ignores the second argument in the put functions.
Because the two formats are fully synchronized, no knowledge of the prior data
is needed.
